# Session 3

In this session we will learn:

* lists, another data structure
* S4 objects
* the DGEList object from the package limma

## Lists

Recall that we have previously learnt about the data structures: vectors, matrices and dataframes. Another important data structure is the list. Like a vector, it is 1 dimensional i.e. one row of data. Unlike vectors, you can put several data types in a list. Here, our list includes data of the integer, a character and a double types:

```{r}
list(1, "a", 1.5)
```


Not only can you put different data types into a list, you can also put a WHOLE data structure into one element in a list. In the list below, the first element is a vector that contains 3 numbers, the second element is a character and the third element is a dataframe that has two columns.


```{r}
list(c(1,2,3), 
     c("words", "letters"), 
     data.frame(column1 = c(1,2,3), column2 = c("a","b","c"))
     )
```

The output can often help you understand how the list is structured. The double brackets (e.g. `[[1]]`) signifies an element of the list and which index it is at. Here there are three elements in our list so the numbers in the double square brackets go from 1 to 3. Underneath the `[[1]]` and `[[2]]`, there is a `[1]` - this indicates that the first and second elements both contain a vector. Underneath `[[3]]` you see the standard output for a dataframe, which we have seen before.

You even include a list within a list (within a list, within a list....I call this 'listception'). This is where it starts to get confusing.

```{r}
list(1, list(1,"a"))
```

* The first element, indicated by `[[1]]`, is a vector, indicated by the `[1]` underneath.
* The second element, indicated by the first `[[2]]` contains a list:
    * `[[2]][[1]]` - tells you that the second element is a list, of which the first element of the inner list is the number 1.
    * `[[2]][[2]]` - tells you that the second element is a list, of which the second element of the inner list is "a".
    
Many bioconductor packages, including limma, use lists so it is an important data structure to understand.

**Challenge 3.1**

Below is the output from a list. 

1. How many element of the list are there?

2. Look carefully at each element and answer the following questions about EVERY element of the list:

  * How many elements does the element of the list contain? 
  * If there are several elements within this element, what does each element contain?


```{r, echo=FALSE}
list(
  list(c(1,2),  c("b")), 
  c("a","b"),
  list(list("a", 1), "b"),
  "end"
  )
```


## S4 objects

Everything in R is an 'object' - every function that we have used and every data structure created in R. All the objects we have used so far, are 'S3' objects. 

Many data structures in Bioconductor packages are 'S4' objects. 'S4' objects are much more 'formal' than S3 objects. 

Let's look at an example. The `DGEList` object from limma is an S4 object. It is a data structure specifically designed for storing read count data from sequencing. It is a list that contains two specific components:

* `counts` - which must be a numeric matrix, that stores counts. Each row must be a gene and each column must be a sample.
* `samples` - which must be a dataframe, that contains information about each sample. Each row must be a sample and must contain information about the group (e.g. treatment group) the sample belongs to, the library size of that sample and the normalisation factor for that sample.

There are a lot more specifications about the type of and how data is stored in the `DGEList` S4 object, than say the S3 list. You can think of S3 objects as very genetic tools (e.g. a spanner) whereas S4 objects tend to be tools designed for a very specific task (e.g. a lug wrench, which is used to loosen and tighten lug nuts on car wheels).


You can read more about the S4 objects in [this](https://www.bioconductor.org/help/course-materials/2017/Zurich/S4-classes-and-methods.html#strengths-of-s4-compared-to-s3) Bioconductor tutorial.


## DGEList


## Answers

**Challenge 3.1**

1. There are 4 elements of this list.
2.

* Within the first element, there are 2 elements. The first is a vector containing 2 numbers and the second is a vector containing one character type (note there are no 'scalars' in R, thus `"a"` is a vector with 1 element). 
* The second element contains 1 element. It is a vector containing two character types.
* The third element contains 2 elements. Within the first element is another list. Within this list there are 2 elements, both being character types. The second element of this nested list is a vector containing one character type.
* The fourth element is a vector containing one character type.









